---
title: "Problem Set 4"
subtitle: "Quantitative Political Science"
author: "[YOUR NAME]"
institute: "Vanderbilt University"
date: "Due Date: 2024-10-28"
output:
  html_document: default
---

```{css,echo = F}
.small .remark-code { /*Change made here*/
  font-size: 85% !important;
}
.tiny .remark-code { /*Change made here*/
  font-size: 50% !important;
}
```

```{r,include=F}
set.seed(123)
options(width=60)
knitr::opts_chunk$set(fig.align='center',fig.width=9,fig.height=5)
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
require(tidyverse)
require(ggpubr)
```

# Instructions


For this assignment, you are to use the `cpsnov2012.dta` dataset found on our class github. This is the November 2012 Current Population Survey, conducted by the U.S. Census Bureau with a nationally representative sample of U.S. households. The codebook for the dataset ([`cpsnov2012.pdf`](https://www2.census.gov/programs-surveys/cps/techdocs/cpsmar12.pdf)) may also be found on our class github.

*Note:* Most of these questions require that you analyze the CPS's household income variable (`hefaminc`). It is coded at the ordinal level, but the analyses require that it be at an interval level. To do this, create a recoded version of `hefaminc` in which each case is assigned a household income equal to the midpoint of its interval of `hefaminc`. For example, a household with income in the range of $5,000-$7,500 should be assigned the value $6,250, and so on.


---

# Question 1

The following scatterplot was produced by an analyst interested in exploring the relationship between turnout (measured with variable `pes1`) and household income (`hefaminc`).

```{r,message=F,warning=F}
require(tidyverse)
require(haven)

dat <- read_dta('./cpsnov2012.dta')

dat %>%
  ggplot(aes(x = pes1,y = hefaminc)) + 
  geom_point()

```

  a. There are many, many things wrong with this figure with regard to both accuracy and style. Name as many as you can.

> Write answer here

  b. Construct a well-designed figure that best displays the relationship between household income and turnout. This will require recoding variables and thinking carefully about the levels at which both variables are measured. Provide the `R` commands you used to recode variables and construct the figure. In a brief paragraph, explain why you made the choices that you did.

```{r}
# INSERT CODE HERE
```

  c. In a few sentences, describe the relationship you see between income and turnout.

> Write answer here

# Question 2

Which relationship--that between income and turnout, or education and turnout--best approximates a linear relationship? The variable to use for educational attainment is `peeduca`. Note that it, like `hefaminc`, is coded at the ordinal level but you want to analyze it as an interval-level variable.

```{r}
# INSERT CODE HERE
```

> Write answer here

# Question 3

You want to investigate the relationship between country of birth and current income.

You wish to divide the sample into three groups: (1) those not born in the U.S.; (2) those born in the U.S. but who have at least one parent not born in the U.S.; and (3) those born in the U.S. with both parents born in the U.S. Using the variables `hefaminc`, `penatvty`, `pemntvty`, and `pefntvty`, construct a boxplot which displays the distribution of household income for each of these three groups.

*Hint:* doing this will require creating new variables from `penatvty`, `pemntvty`, and `pefntvty`. You will also need to make choices about how to deal with missing values. Justify any choices you make in a note accompanying the figure.

```{r}
# INSERT CODE HERE
```

> Write answer here

Using the proper statistical tests with an $\alpha =.05$, answer the following questions. Note that additional recoding may be necessary.

  a. Do native-born Americans have higher incomes than non-native born Americans?

```{r}
# INSERT CODE HERE
```

> Write answer here

  b. Do native-born Americans whose parents were born in the U.S. have higher incomes than native-born Americans with at least one parent **not** born in the U.S.?
```{r}
# INSERT CODE HERE
```

> Write answer here

  c. Do Americans with a foreign-born father and a native-born mother have lower incomes than Americans with a foreign-born mother and a native-born father?

```{r}
# INSERT CODE HERE
```

> Write answer here

In a few sentences, describe your results.

> Write answer here

