---
title: "Problem Set 8"
subtitle: "Quantitative Political Science"
author: "[YOUR NAME]"
institute: "Vanderbilt University"
date: "Due Date: 2024-11-04"
output:
  html_document: default
---

```{css,echo = F}
.small .remark-code { /*Change made here*/
  font-size: 85% !important;
}
.tiny .remark-code { /*Change made here*/
  font-size: 50% !important;
}
```

```{r,include=F}
set.seed(123)
options(width=60)
knitr::opts_chunk$set(fig.align='center',fig.width=9,fig.height=5)
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
require(tidyverse)
require(ggpubr)
```

# Instructions


For this assignment, we are going to use a new dataset on the education industry in the United States from 2019 called `sc_debt.Rds`. You can find this on the course Github at [https://github.com/jbisbee1/PSCI_8356_F2024/raw/refs/heads/main/Data/sc_debt.Rds](https://github.com/jbisbee1/PSCI_8356_F2024/raw/refs/heads/main/Data/sc_debt.Rds).

The codebook is reproduced here.

| Name           | Definition                                                                                                                                                                                             |
|----------------|--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| unitid         | Unit ID                                                                                                                                                                                                |
| instnm         | Institution Name                                                                                                                                                                                       |
| stabbr         | State Abbreviation                                                                                                                                                                                     |
| grad_debt_mdn  | Median Debt of Graduates                                                                                                                                                                               |
| control        | Control Public or Private                                                                                                                                                                              |
| region         | Census Region                                                                                                                                                                                          |
| preddeg        | Predominant Degree Offered: Associates or Bachelors                                                                                                                                                    |
| openadmp       | Open Admissions Policy: 1= Yes, 2=No,3=No 1st time students                                                                                                                                            |
| adm_rate       | Admissions Rate: proportion of applications accepted                                                                                                                                                   |
| ccbasic        | Type of institution-- see [here](https://data.ed.gov/dataset/9dc70e6b-8426-4d71-b9d5-70ce6094a3f4/resource/658b5b83-ac9f-4e41-913e-9ba9411d7967/download/collegescorecarddatadictionary_01192021.xlsx) |
| selective      | Institution admits fewer than 10 % of applicants, 1=Yes, 0=No                                                                                                                                          |
| research_u     | Institution is a research university 1=Yes, 0=No                                                                                                                                                       |
| sat_avg        | Average Sat Scores                                                                                                                                                                                     |
| md_earn_wne_p6 | Average Earnings of Recent Graduates                                                                                                                                                                   |
| ugds           | Number of undergraduates                           |
| costt4a        | Average cost of attendance (tuition-grants)        |

---


# Question 1

Let's start by looking at the relationship between an ordinal variable and a ratio variable. Start by looking through the dataset and choosing one of each (note you cannot use the same two variables that were used in the lecture notes for any of these). What do you think the relationship between these two variables should be? Why? Based on this answer, which variable should be your $X$ and which variable should be your $Y$?

> Write answer here.

```{r}
require(tidyverse)
dat <- read_rds("https://github.com/jbisbee1/PSCI_8356_F2024/raw/refs/heads/main/Data/sc_debt.Rds")

dat
```

  a. First, create univariate visualizations of both variables.

```{r}
# INSERT CODE HERE
```

  
  b. Second, create a multivariate visualization comparing one variable to the other.

```{r}
# INSERT CODE HERE
```


# Question 2

Now let's choose two ratio variables. What do you think the relationship between these two variables should be? Why? Based on this answer, which variable should be your $X$ and which variable should be your $Y$?

> Write answer here

  a.  As before, describe the relationship between these two variables non-parametrically. Do so first by calculating the conditional means between your $X$ and $Y$ variables. Then visualize this relationship.
  
```{r}
# INSERT CODE HERE
```
  
  b. Next, describe the relationship between these two variables parametrically. Do so first by manually calculating the correlation coefficient based on the raw math from class (i.e. $\frac{\sum_i(X_i - \bar{X})(Y_i - \bar{Y})}{\sqrt{\sum_i (X_i - \bar{X})^2 \sum_i (Y_i - \bar{Y})^2}}$). Then use the `R` functions `cov()`, `sqrt()` and `var()` to redo the calculation and see if you get the same answer. Finally, use the `R` function `cor()` to calculate the same number again. Do you get the same value across all three cases? If not, why not?
  
```{r}
# INSERT CODE HERE
```

  c. Calculate the linear regression of the form $y_i = \beta_0 + \beta_1 x_1 + \varepsilon$ in `R` using the `lm()` function. Write a short description of the results, making sure to express the coefficients appropriately given the units of the variables you've chosen.
  
  d. Now recalculate $\hat{\beta}_0$ manually by using the equation on slide 23 of Lecture 16. Do the same for $\hat{\beta}_1$.
  
  e. Re-do the manual calculation for $\hat{\beta}_1$ using the equation found on slide 26 of Lecture 16. What does this remind you of?
  
  f. Re-do the manual calculation for $\hat{\beta}_0$ using the equation found on slide 27 of Lecture 16.
  
# Question 3

A few fun maths for old time's sake!

  a. Prove that $SST = SSR + SSE$
  
  b. Solve

$$
\mathbf{A}=
\begin{bmatrix}
2 & 0 \\ 
-5 & 3
\end{bmatrix}
\text{ and }
\mathbf{B=}
\begin{bmatrix}
4 & 10 \\ 
1 & 3
\end{bmatrix}
$$

$$
\mathbf{AB}=
\begin{bmatrix}
? & ? \\ 
? & ?
\end{bmatrix}
$$

  c. Solve
  
$$
\begin{bmatrix}
n & \sum_ix_i \\ 
\sum_ix_i & \sum_ix_i^2
\end{bmatrix}
^{-1}
= ~~
?
$$