---
title: "Problem Set 6"
author: '[YOUR NAME]'
date: "Due Date: 2024-11-18"
output:
  html_document: default
  pdf_document: default
subtitle: Quantitative Political Science
institute: Vanderbilt University
---

```{css,echo = F}
.small .remark-code { /*Change made here*/
  font-size: 85% !important;
}
.tiny .remark-code { /*Change made here*/
  font-size: 50% !important;
}
```

```{r,include=F}
set.seed(123)
options(width=60)
knitr::opts_chunk$set(fig.align='center',fig.width=9,fig.height=5)
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
require(tidyverse)
require(ggpubr)
```

# Instructions


For this assignment, you are to use the `counties.dta` dataset found on our class [github](https://github.com/jbisbee1/PSCI_8356_F2024/raw/refs/heads/main/Data/counties.dta). It comes from the City and County Data Book produced by the U.S. Census Bureau. 



```{r,message=F,warning=F}
require(tidyverse)
require(haven)

dat <- read_dta('https://github.com/jbisbee1/PSCI_8356_F2024/raw/refs/heads/main/Data/counties.dta')
```

#### 1: Analyze the following questions using `R`, but answer them with a few sentences in plain English.

a. Controlling for household income (`hhincome`), is the percentage of a county’s residents who have college degrees (`college`) associated with crime rates (`crimerate`)? How so?

b. Construct a plot displaying the relationship between $\widehat{\textit{crimerate}}$ and `college`. To do this, write a loop that generates the value of $\widehat{\textit{crimerate}}$ when `college` is equal to $5, 7,\dots,25$. (In doing so, you may find it helpful to use the coefficients.) Then plot these connected points.

c. You are interested in how well this model applies to counties in New York State. Construct a plot showing at a glance that Tompkins County, has a crime rate much lower than predicted by a regression of `crimerate` on `college` among counties in New York State.

d. Look up Tompkins County on the Internet. Why might it be an outlier? Does this suggest an additional control variable?

#### 2: Is unemployment higher in counties located in the South than those located outside the South?

a. Answer this question first with a $t$-test to compare group means.

b. Now answer this question by running a bivariate specification regressing `unemprate` on the indicator variable `south`.

c. Identify two important similarities found between the two analyses.

d. We require a lot more assumptions to perform the second analysis than the first. In a few sentences, reflect upon why they are unnecessary to answer the question posed above.

#### 3: Is the association between poverty and crime by county stronger in the South or outside the South?

a. Answer this question first by running two separate regressions.

b. Now answer this question using an interacted specification.

c. Identify two important similarities found between the estimated generated by the two analyses.

d. Construct a plot with two lines: one showing $\widehat{\textit{crimerate}}$ by `povrate` among counties in the South, and the other showing $\widehat{\textit{crimerate}}$ by `povrate` among non-Southern counties.

#### 4: Perform the necessary analyses to assess the accuracy of the following statements. In a few sentences, discuss each of your findings.

a. The bivariate relationship between poverty and crime is weaker where local government spending (`locspendcap`) is higher.

b. The bivariate relationship between unemployment and crime is stronger in densely populated counties (`density`).

c. Federal spending per capita (`fedspendcap`) plays a greater role in reducing poverty than local spending per capita (`locspendcap`), controlling for variables that may confound these relationships.

#### 5: Pretend to **not** be a dispassionate scholar but instead are someone with an agenda who wants to show that crime is caused by poverty. To do so, you are hell-bent on finding the combination of regressors (including `povrate`) that results in a very large coefficient on `povrate` when `crimerate` is regressed upon these regressors.

a. Choose 90 percent of your observations at random and set them aside. Do not include these observations in the analysis.

b. Find the combination of variables that yields the highest possible coefficient on `povrate` as a predictor of `crimerate` that you can obtain.

c. Once you've arrived at a specification, note the results. Now, take the rest of your dataset out of "cold storage." How well does your model do with these other data? What is the lesson learned?
