---
title: "Review Session 2"
subtitle: "Quantitative Political Science"
author: "Prof. Bisbee"
institute: "Vanderbilt University"
date: "Lecture Date: 2024/09/19\n Slides Updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    # self_contained: true
    chakra: libs/remark-latest.min.js
    lib_dir: libs
    css:
      - default
      - css/lexis.css
      - css/lexis-fonts.css
    #seal: false
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"

---

```{css,echo = F}
.small .remark-code { /*Change made here*/
  font-size: 85% !important;
}
.tiny .remark-code { /*Change made here*/
  font-size: 50% !important;
}
```

```{r,include=F}
set.seed(123)
options(width=60)
knitr::opts_chunk$set(fig.align='center',fig.width=9,fig.height=5)
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
require(tidyverse)
```

# Agenda

2. Calculating areas under curves

1. Probability distributions

3. Putting the two together

4. The Normal

---

# Areas under curves

- Let's draw some generic function $y = 100 - X^2$

```{r}
p <- data.frame(X = -10:10) %>%
  mutate(Y = 100 - X^2) %>%
  ggplot(aes(x = X,y = Y)) + 
  # geom_bar(stat = 'identity')
  geom_line()
```

---

# Areas under curves

- Let's draw some generic function $y = 100 - X^2$

```{r}
p
```

---

# Areas under curves

- What is the total area under the curve between $-10 \leq X \leq 10$?

- We could divide this up into 20 bars

```{r}
p + 
  geom_bar(stat = 'identity',color = 'black',fill = 'white',alpha = .3)
```

---

# Areas under curves

- These are just a bunch of rectangles, right?

- The base of each is 1, and the height is $100 - X^2$

--

  - Thus the first rectangle at $X = -10$ is $1*(100 - (-10^2)) = 1*(100 - 100) = 1*0 = 0$
  
  - The second rectangle at $X = -9$ is $1*(100 - (-9^2)) = 1*(100 - 81) = 1*19 = 19$
  
--

- Just keep on calculating each rectangle and then add them all up when done!

---

# Areas under curves

- But the preceding approach makes big mistakes...the rectangles don't fit perfectly under the curve!

- Solution? Make the rectangles less wide!

```{r,echo=F}
data.frame(X = seq(-10,10,by = 1)) %>%
  mutate(Y = 100 - X^2) %>%
  ggplot(aes(x = X,y = Y)) + 
  # geom_bar(stat = 'identity')
  geom_bar(stat = 'identity',color = 'black',fill = 'white',alpha = .3) + 
  geom_line()
```

---

# Areas under curves

- But the preceding approach makes big mistakes...the rectangles don't fit perfectly under the curve!

- Solution? Make the rectangles less wide!

```{r,echo=F}
data.frame(X = seq(-10,10,by = .5)) %>%
  mutate(Y = 100 - X^2) %>%
  ggplot(aes(x = X,y = Y)) + 
  # geom_bar(stat = 'identity')
  geom_bar(stat = 'identity',color = 'black',fill = 'white',alpha = .3) + 
  geom_line()
```

---

# Areas under curves

- But the preceding approach makes big mistakes...the rectangles don't fit perfectly under the curve!

- Solution? Make the rectangles less wide!

```{r,echo=F}
data.frame(X = seq(-10,10,by = .25)) %>%
  mutate(Y = 100 - X^2) %>%
  ggplot(aes(x = X,y = Y)) + 
  # geom_bar(stat = 'identity')
  geom_bar(stat = 'identity',color = 'black',fill = 'white',alpha = .3) + 
  geom_line()
```

---

# Areas under curves

- But the preceding approach makes big mistakes...the rectangles don't fit perfectly under the curve!

- Solution? Make the rectangles less wide!

```{r,echo=F}
data.frame(X = seq(-10,10,by = .1)) %>%
  mutate(Y = 100 - X^2) %>%
  ggplot(aes(x = X,y = Y)) + 
  # geom_bar(stat = 'identity')
  geom_bar(stat = 'identity',color = 'black',fill = 'white',alpha = .3) + 
  geom_line()
```

---

# Areas under curves

- We can keep on narrowing the widths until they infinitesimally small

- In math, we call this "integration" and denote it $\int y dy$

--

  - Note that $y$ is the output of our function $f(x) = y = 100 - X^2$, so we can instead write $\int f(x) dx$
  
  - And $dx$ literally represents narrowing the width of $X$ to a vanishingly small value

--

- To calculate this, following the three-step process

  1. Figure out the **anti-derivative** of our function
  
  2. Evaluate this anti-derivative at the maximum value and the minimum values we want to integrate over
  
  3. Subtract the value at the minimum from the value at the maximum
  
---

# Areas under curves

$$
\begin{aligned}
\int_{-5}^5 f(x)dx &= \int_{-5}^5 (100 - X^2) dx\\
&= 100X-\frac{X^3}{3} \bigg|_{-5}^5\\
&= 100*5 - \frac{5^3}{3} - \bigg(100*-5 - \frac{-5^3}{3}\bigg)\\
&= 458.33 - -458.33\\
&= 916.67
\end{aligned}
$$

---

# Areas under curves

```{r,echo=F}
data.frame(X = seq(-10,10,by = .1)) %>%
  mutate(Y = 100 - X^2) %>%
  mutate(fl = X > -5 & X < 5) %>%
  ggplot(aes(x = X,y = Y,color = fl)) + 
  geom_line() + 
  geom_bar(stat = 'identity') + 
  theme(legend.position = 'none')
```

  
---

# Probability Distributions

- A discrete random variable $Y$ can take on either:

--

  - A countably infinite number of values $y$
  
  - A finite number of values $y$
  
--

- Each of these values $y$ has some probability $p(y)$ of appearing

---

# Probability Distributions

- Imagine a random variable that can take on the values 1, 2, or 3

- Assume further that the probability of $Y = 1$ is 0.3, $Y = 2$ is 0.5, and $Y = 3$ is 0.2

--

- This is a discrete random variable with a probability distribution of $0.3, 0.5, 0.2$

---

# Probability Distributions

- Let's draw some observations from it!

```{r,warning=F,message=F}
require(tidyverse)
sample(1:3,replace = T,prob = c(.3,.5,.2))
sample(1:3,replace = T,prob = c(.3,.5,.2))
```

---

# Probability Distributions

- Let's draw 10,000 observations!

```{r}
Y <- sample(1:3,size = 10000,replace = T,prob = c(.3,.5,.2))
data.frame(Y = Y) %>%
  as_tibble()
```

---

# Probability Distributions

- Let's now count up the number of times each $y$ outcome is realized, and calculate their proportions
```{r}
df <- data.frame(Y = Y) %>%
  as_tibble() %>%
  count(Y) %>%
  mutate(p_y = n / sum(n))
```

--

- Not exactly right, but pretty close

---

# Probability Distributions

- This vector of $p(y)$ is the **probability distribution** for this random variable

- We can visualize it

```{r}
p <- df %>%
  ggplot(aes(x = Y,y = p_y)) + 
  geom_bar(stat = 'identity')
```

---

# Probability Distributions

```{r}
p
```


---

# Expectations

- Recall the expectation of a discrete random variable?

--

- $E[Y] = \sum y p(y)$

--

```{r}
sum(1*.3,2*.5,3*.2)
```

---

# Expectations

- Could also calculate it from our sample of 10,000 observations

```{r}
df %>%
  mutate(y_X_py = Y*p_y) %>%
  summarise(E_Y = sum(y_X_py))
```

---

# Probabilities

- What if we wanted to know the probability that $Y = 3$?

--

- Just look it up! $p(Y=3) = 0.2$

--

- What if we wanted to know the probability that $Y < 3$?

--

- $P(Y=1 \cup Y=2) = ?$

--

- Recall that random variables take on values that are **mutually exclusive**

$$
\begin{aligned}
P(Y=1 \cup Y=2) &= P(Y=1) + P(Y=2)\\
&= 0.3 + 0.5\\
&= 0.8
\end{aligned}
$$

---

# Probability Distributions

- For continuous random variables, $p(Y=y) = 0$

--

```{r}
Y <- rgamma(1000,shape = 1)
(df <- data.frame(Y = Y) %>%
  as_tibble() %>%
  count(Y) %>%
  mutate(p_y = n / sum(n)))
```

---

# Probability Distributions

- But $p(Y=y)$ is not zero!

--

- Sure, but this is because I have drawn a **finite sample**. If I increase $n$, $p(Y=y)$ will get smaller

```{r}
Y <- rbeta(n = 100000,2,3)
(df <- data.frame(Y = Y) %>%
  as_tibble() %>%
  count(Y) %>%
  mutate(p_y = n / sum(n)))
```

---

# Probability Distributions

- However, even though any individual $P(Y = y) = 0$, this distribution still has a shape

--

- We can **bin** the random variable and then count the number of observations in bins

```{r,message=F,echo=F}
df %>%
  ggplot(aes(x = Y)) + 
  geom_histogram()
```


---

# Probability Distributions

- But what "shape" does this binning reveal? It depends on the number of bins!

```{r}
df %>%
  ggplot(aes(x = Y)) + 
  geom_histogram(bins = 10)
```

---

# Probability Distributions

- But what "shape" does this binning reveal? It depends on the number of bins!

```{r}
df %>%
  ggplot(aes(x = Y)) + 
  geom_histogram(bins = 100)
```

---

# Probability Distributions

- But what "shape" does this binning reveal? It depends on the number of bins!

```{r}
df %>%
  ggplot(aes(x = Y)) + 
  geom_histogram(bins = 20)
```

---

# Probability Distributions

- Nevertheless, we would like to imagine that we can define a **function** which will draw this shape

--

  - This function is called the probability distribution function, or PDF, and is denoted $f(y)$
  
  - $f(y) \geq 0~ \forall ~y: -\infty < y < \infty$
  
  - $\int_{-\infty}^\infty f(y)dy = 1$

---

# Probability Distributions

- Since the area under the curve of a PDF has to sum to 1 (axioms!), subsets are probabilities

```{r,echo = F}
df %>%
  mutate(Y = round(Y*30,0)/30) %>%
  group_by(Y) %>%
  summarise(p_y = sum(p_y)) %>%
  ggplot(aes(x = Y,y = p_y)) + 
  geom_bar(stat = 'identity',color = 'black',alpha = .3) + 
  geom_line()
```


--

- I.e., What is the probability that the random variable is between 0.1 and 0.3?

--

- Integrate! Except...what is the function?

---

# Probability Distributions

- "Easy" functions include Bernoulli, Binomial, and Poisson (discrete) and the uniform (continuous)

- But we wind up spending most of our time with the Normal

- Why?

--

- Basic answer: "it appears in nature a lot!"

- Galaxy brain answer: "ALL DISTRIBUTIONS BECOME NORMAL VIA THE CENTRAL LIMIT THEOREM!"

---

# Central Limit Theorem

```{r,echo = F,message=F}
set.seed(123)
Y <- rexp(1000000)
data.frame(Y = Y) %>%
  as_tibble() %>%
  ggplot(aes(x = Y)) + 
  geom_histogram()
```

---

# Central Limit Theorem

```{r,echo = F}
mean(sample(Y,size = 1000))
```

```{r,echo = F}
mean(sample(Y,size = 1000))
```

```{r,echo = F}
mean(sample(Y,size = 1000))
```

--

- Do this many times!

```{r}
means <- NULL
for(i in 1:1000) {
  means <- c(means,mean(sample(Y,size = 1000)))
}
```

---

# Central Limit Theorem

- What does the distribution of the repeatedly randomly sampled means look like? **THE NORMAL BABY!**

```{r,echo=F,message=F}
data.frame(means = means) %>%
  ggplot(aes(x = means)) + 
  geom_histogram()
```


---

# The Normal

- We won't get into the math of the normal, but it is an essential part of quantitative analysis!

- SO just trust me:

  - $f(y) = \frac{1}{\sigma \sqrt{2\pi}}e^{\frac{-(y-\mu)^2}{2\sigma^2}}$ for $-\infty < y < \infty$
  
- Two parameters: $\mu$ and $\sigma$

  - $E(Y) = \mu$
  
  - $\textit{VAR}(Y) = \sigma^2$
  
---

# The Normal
  
- What is the probability $Y$ takes on some value $y$ within an interval between $a = 62$ and $b = 65$?

  - $P(a \leq Y \leq b) = \int_{a}^b\frac{1}{\sigma\sqrt{2\pi}}e^{\frac{-(y-\mu)^2}{2\sigma^2}}dy$

--

```{r,echo=F,message=F,warning=F}
require(tidyverse)
data.frame(y = seq(50,75,length.out = 1000)) %>%
  mutate(py = dnorm(y,mean = 63,sd = 3)) %>%
  mutate(area = y >= 62 & y <= 65) %>%
  ggplot(aes(x = y,ymin = 0,ymax = py,fill = area)) + 
  geom_ribbon() +
  geom_ribbon(fill = NA,color = 'black') + 
  scale_fill_manual(guide = 'none',values = c('white','grey70')) + 
  theme_bw() + 
  geom_vline(xintercept = c(62,65),linetype = 'dashed')
```

---

# The Normal

- We typically **standardize** a normally distributed variable

--

  - Units measured in terms of standard deviations (instead of inches or whatever else)
  
- $Z \equiv \frac{Y - \mu}{\sigma}$

  - $Z$ is a random variable with mean zero and standard deviation one
  
  - PDF simplifies to $f(z) = \frac{1}{\sqrt{2\pi}}e^{-\frac{z^2}{2}}$
  
- We use these so frequently in statistics we denote them with special symbols!

  - "Little phi of z" is the PDF of the standardized normal evaluated at $Z = z$: $\phi(z)$
  
  - "Big phi of z" is the CDF of the standardized normal evaluated at $Z = z$: $\Phi(z)$
  
---

# The Normal

- `R` has some handy functions that we can use to calculate:

--

  1. The probability that a standard Normal is less than some number of standard deviations: `pnorm()`
  
  2. The number of standard deviations that enclose some probability of a standard Normal: `rnorm()`
  
---

# The Normal

- Consider a Normal with mean = 63 and standard deviation = 3

- In math: $\mathcal{N}(\mu,\sigma) = \mathcal{N}(63,3)$

```{r,echo=F,message=F,warning=F}
require(tidyverse)
data.frame(y = seq(50,75,length.out = 1000)) %>%
  mutate(py = dnorm(y,mean = 63,sd = 3)) %>%
  mutate(area = y >= 62 & y <= 65) %>%
  ggplot(aes(x = y,ymin = 0,ymax = py)) + 
  # geom_ribbon() +
  geom_ribbon(fill = NA,color = 'black') + 
  scale_fill_manual(guide = 'none',values = c('white','grey70')) + 
  theme_bw() + 
  labs(y = 'f(y)') + 
  geom_vline(xintercept = 63,linetype = 'dashed') + 
  geom_vline(xintercept = c(60,66),linetype = 'dotted')
  # geom_vline(xintercept = c(62,65),linetype = 'dashed')
```

---

# The Normal

- What is the probability that $Y \leq 66$?

--

```{r}
pnorm(1)
```

--

```{r,echo=F,message=F,warning=F}
require(tidyverse)
data.frame(y = seq(50,75,length.out = 1000)) %>%
  mutate(py = dnorm(y,mean = 63,sd = 3)) %>%
  mutate(area = y <= 66) %>%
  ggplot(aes(x = y,ymin = 0,ymax = py,fill = area)) + 
  geom_ribbon() +
  geom_ribbon(fill = NA,color = 'black') + 
  scale_fill_manual(guide = 'none',values = c('white','grey70')) + 
  theme_bw() + 
  labs(y = 'f(y)') + 
  # geom_vline(xintercept = 63,linetype = 'dashed') + 
  # geom_vline(xintercept = c(60,66),linetype = 'dotted')
  geom_vline(xintercept = c(66),linetype = 'dashed') + 
  annotate(geom = 'label',x = 63,y = .05,label = '84%')
```


---

# The Normal

- What is the probability that $60 \leq Y \leq 66$?

--

```{r}
pnorm(1) - pnorm(-1)
```

--

```{r,echo=F,message=F,warning=F}
require(tidyverse)
data.frame(y = seq(50,75,length.out = 1000)) %>%
  mutate(py = dnorm(y,mean = 63,sd = 3)) %>%
  mutate(area = y >= 60 & y <= 66) %>%
  ggplot(aes(x = y,ymin = 0,ymax = py,fill = area)) + 
  geom_ribbon() +
  geom_ribbon(fill = NA,color = 'black') + 
  scale_fill_manual(guide = 'none',values = c('white','grey70')) + 
  theme_bw() + 
    labs(y = 'f(y)') + 
  # geom_vline(xintercept = 63,linetype = 'dashed') + 
  # geom_vline(xintercept = c(60,66),linetype = 'dotted')
  geom_vline(xintercept = c(60,66),linetype = 'dashed') + 
  annotate(geom = 'label',x = 63,y = .05,label = '68%')
```

---

# The Normal

- What is the probability that $Y \leq 63$?

--

```{r}
pnorm(0)
```

--

```{r,echo=F,message=F,warning=F}
require(tidyverse)
data.frame(y = seq(50,75,length.out = 1000)) %>%
  mutate(py = dnorm(y,mean = 63,sd = 3)) %>%
  mutate(area = y <= 63) %>%
  ggplot(aes(x = y,ymin = 0,ymax = py,fill = area)) + 
  geom_ribbon() +
  geom_ribbon(fill = NA,color = 'black') + 
  scale_fill_manual(guide = 'none',values = c('white','grey70')) + 
  theme_bw() + 
  labs(y = 'f(y)') + 
  # geom_vline(xintercept = 63,linetype = 'dashed') + 
  # geom_vline(xintercept = c(60,66),linetype = 'dotted')
  geom_vline(xintercept = c(63),linetype = 'dashed') + 
  annotate(geom = 'label',x = 61,y = .05,label = '50%')
```


---

# The Normal

- What is the value of $y$ below which 60% of the data is contained?

--

```{r}
qnorm(.6)
```

--

- Think about this for a second! This is **in terms of the standardized Normal**

- This literally means "0.25335 standard deviations away from the mean"

- What is our standard deviation?

--

- $\sigma = 3$

- What is our mean?

--

- $\mu = 63$

--

- (Does "away" mean above or below?)


---

# The Normal

- So to translate back to the unstandardized Normal random variable?

$$
\begin{aligned}
Z &= \frac{Y - \mu}{\sigma}\\
\sigma*Z &= Y - \mu\\
\sigma*Z + \mu &= Y\\
Y &= 3*0.25335 + 63\\
&= 63.76
\end{aligned}
$$

---

# The Normal


```{r,echo=F,message=F,warning=F}
require(tidyverse)
data.frame(y = seq(50,75,length.out = 1000)) %>%
  mutate(py = dnorm(y,mean = 63,sd = 3)) %>%
  mutate(area = y <= 63+(.2533471*3)) %>%
  ggplot(aes(x = y,ymin = 0,ymax = py,fill = area)) + 
  geom_ribbon() +
  geom_ribbon(fill = NA,color = 'black') + 
  scale_fill_manual(guide = 'none',values = c('white','grey70')) + 
  theme_bw() + 
  labs(y = 'f(y)') + 
  # geom_vline(xintercept = 63,linetype = 'dashed') + 
  # geom_vline(xintercept = c(60,66),linetype = 'dotted')
  geom_vline(xintercept = c(63+(.2533471*3)),linetype = 'dashed') + 
  annotate(geom = 'label',x = 62,y = .05,label = '60%')
```

---

# The Normal

- What values of $y$ contain 95% of the data?

--

```{r}
qnorm(.975)
```

--

```{r,echo=F,message=F,warning=F}
require(tidyverse)
data.frame(y = seq(50,75,length.out = 1000)) %>%
  mutate(py = dnorm(y,mean = 63,sd = 3)) %>%
  mutate(area = y >= 63-(1.96*3) & y <= 63+(1.96*3)) %>%
  ggplot(aes(x = y,ymin = 0,ymax = py,fill = area)) + 
  geom_ribbon() +
  geom_ribbon(fill = NA,color = 'black') + 
  scale_fill_manual(guide = 'none',values = c('white','grey70')) + 
  theme_bw() + 
  labs(y = 'f(y)') + 
  # geom_vline(xintercept = 63,linetype = 'dashed') + 
  # geom_vline(xintercept = c(60,66),linetype = 'dotted')
  geom_vline(xintercept = c(63-(1.96*3),63+(1.96*3)),linetype = 'dashed') + 
  annotate(geom = 'label',x = 63,y = .05,label = '95%')
```