---
title: "Odds and Ends"
subtitle: ""
author: "Prof. Bisbee"
institute: "Vanderbilt University"
date: "Lecture Date: 2023/11/30\n Slides Updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    # self_contained: true
    chakra: libs/remark-latest.min.js
    lib_dir: libs
    css:
      - default
      - css/lexis.css
      - css/lexis-fonts.css
    #seal: false
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "17:9"

---

```{css,echo = F}
.small .remark-code { /*Change made here*/
  font-size: 85% !important;
}
.tiny .remark-code { /*Change made here*/
  font-size: 50% !important;
}
```

```{r,include=F}
set.seed(123)
options(width=60)
knitr::opts_chunk$set(fig.align='center',fig.width=9,fig.height=5)
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
require(tidyverse)
require(ggpubr)
```

# Agenda

1. Regression Diagnostics

2. Goodness of Fit

3. Overspecification

4. 

---

# Regression Diagnostics

- We have so many assumptions at this point! (How many can you list?)

--

  1. Linearity
  
  2. I.i.d. random sample
  
  3. Non-zero variance / no multicolinearity
  
  4. Zero-conditional mean
  
  5. Homoskedasticity / spherical errors
  
  6. Normally distributed errors (small samples)
  
--

- How can we be confident in these assumptions? **Diagnostics** (to an extent)

---

# Regression Diagnostics

- Running example of two DGPs

--

- DGP 1:

$$
\begin{aligned}
\textit{Income}_i &= 15 + 6*\textit{Labor}_i + 40*\textit{PhD}_i + u_i
\end{aligned}
$$

where $u \sim \mathcal{N}(0,5)$ and i.i.d. holds

--

- DGP 2:

$$
\begin{aligned}
\textit{Income}_i &= 15 + 6*\textit{Labor}_i + 40*\textit{PhD}_i + \textit{Labor}_i^2 + u_i
\end{aligned}
$$
where $u_i = 0.5*\textit{Labor}_i*e_i$,  $e_i \sim \mathcal{N}(0,5)$ and i.i.d. holds

--

- Note that our assumptions hold by construction in DGP 1, but not in DGP 2

--

  - Specifically, zero conditional mean holds only if *Labor* is mean zero
  
  - In addition, the errors are not homoskedastic
  
---

# Regression Diagnostics

- Estimate with $\textit{Income}_i = \beta_0 + \beta_1 \textit{Labor}_i + \beta_2 \textit{PhD}_i + u_i$

```{r,echo=F,warning=F}
# PROPS TO JML!
set.seed(12345)


################################################################
## This file aims to do three things:
## 1) Demonstrate what is meant by repeated sampling in the regression context
## 2) Make crystal clear what the assumptions required for inference say about
## the data generating process
## 3) Show how to use diagnostic plots to glean that something might be off
## about the assumptions.


################################################################
## First, let's make some data that perfectly comports with the assumptions
## required for inference in OLS


## To start, we need to conjure up our independent variables.  
## Let's make a dummy variable for holding a PhD or not, and an interval
## variable for years in the labor force.

phd <- c(rep(0, 75), rep(1, 75))
labor <- sample(c(1:30), size = 150, replace = TRUE)


## Now let's generate our dependent variable, income, as a linear function of 
## years in labor market and degree status.
## In other words, we're assuming we know the data generating process.
## Let's have our DGP meet the assumptions of inference in OLS, and
## let's use it to generate some data (by which we mean generate the DV, income)

## We'll need a normally-distributed error term with mean 0 and a constant SD
error <- rnorm(150, mean = 0, sd = 5)

## Let's assume the true parameters are:
beta0 <- 15
beta1 <- 6
beta2 <- 40

## So now, let's generate one sample of our DV:

inc <- beta0 + beta1*labor + beta2*phd + error

## This is the step that would be repeated under "repeated sampling."

## Now let's collect our IVs and this DV into a dataframe.

data <- as.data.frame(cbind(inc, labor, phd))

## So we've created our outcome variable as a linear function of labor and phd,
## with normally-distributed error cenetered on zero with constant variance (sd = 5)


## Ok, now let's see what we get when we regress income on labor and phd
## using these data
```

```{r}
m1 <- lm(inc ~ labor + phd, data = data)
## Did we reproduce the truth?
summary(m1)
```

---

# Regression Diagnostics

- Visualize the results

```{r,echo=F,warning=F,fig.width=16,fig.height=6}
## And let's plot the regression lines separately for PhD and not
plot(labor, inc)
abline(coef(m1)[1], coef(m1)[2], lty = 2)
abline(coef(m1)[1]+coef(m1)[3], coef(m1)[2], lty = 1)

legend("topleft", legend = c("PhD", "No PhD"), title = "Degree", lty = c(1,2), lwd = c(1,1), cex = 1)
```

---

# Regression Diagnostics

- What if we are in the DGP2 world?

```{r,echo=F,warning=F}
## Let's contrast this with a true DGP that violates some of the assumptions
## for inference in OLS:

## For comparison, let's make another dataset

## Let's use the same IVs
phd <- c(rep(0, 75), rep(1, 75))
labor <- sample(c(1:30), size = 150, replace = TRUE)

## But notice the differences in the error term we're going to assume:

error <- .5*labor * rnorm(150, mean = 0, sd = 5)

## Let's keep the same assumed true parameter values

beta0 <- 15
beta1 <- 6
beta2 <- 40

## And let's generate our DV in this way:

inc <- beta0 + beta1*labor + beta2*phd + (labor^2) +  error

data2 <- as.data.frame(cbind(inc, labor, phd))
```

```{r}
m2 <- lm(inc ~ labor + phd, data = data2)

summary(m2)
```

---

# Regression Diagnostics

```{r,echo=F,warning=F,fig.height=6,fig.width=16}
## And plot it:

par(mfrow = c(1,1))
plot(labor, inc)
abline(coef(m2)[1], coef(m2)[2], lty = 2)
abline(coef(m2)[1]+coef(m2)[3], coef(m2)[2], lty = 1)

legend("topleft", legend = c("PhD", "No PhD"), title = "Degree", lty = c(1,2), lwd = c(5,5), cex = 1)
# plot(m2)
```

---

# Regression Diagnostics

- We are doing a bad job because:

  1. Income is not a linear function of labor and degree
  
  2. Errors are not mean zero
  
  3. Errors are not homoskedastic
  
--

- We **know** all this because we simulated these data

- But in reality, we typically never know what the true DGP is...how can we be alerted to the fact something is wrong with our model?

---

# Regression Diagnostics

- We can **look** at our residuals in a number of ways that inform us about our model's fit

--

1. Residuals vs. Fitted Values: Tells us if the data are roughly linear (smoother is roughly a horizontal line) and if there is heteroskedasticity (residuals are larger for some observations than for others)

2. Normal Q-Q Plot: Compares quantiles of our observed residuals to quantiles of hypothetical residuals that are normally distributed. If points cling to the 45 degree line, the residuals are normally distributed.

3. Scale-Location Plot: Similar to Residuals vs. Fitted, except we put the y-axis is now the square root of the standardized residuals. Also informs us about heteroskedasticity (shouldn't see a pattern) and identifies potential outliers

4. Residuals vs.k Leverage: Visualizes the **influence** of each observation on the regression coefficients. Points that are far from other points, especially those that are close to the dashed red lines, are problematic.

--

- Let's look at each in turn (`R` will produce all four by default if you simply run `plot(m1)` on your regression model)

---

# Residuals vs. Fitted


```{r,echo=F,warning=F,fig.width=14,fig.height=7}
par(mfrow = c(1,2))
plot(m1,which = 1)
plot(m2,which = 1)
```

---

# Normal Q-Q Plot


```{r,echo=F,warning=F,fig.width=14,fig.height=7}
par(mfrow = c(1,2))
plot(m1,which = 2)
plot(m2,which = 2)
```

---

# Scale-Location


```{r,echo=F,warning=F,fig.width=14,fig.height=7}
par(mfrow = c(1,2))
plot(m1,which = 3)
plot(m2,which = 3)
```

---

# Residuals vs. Leverage


```{r,echo=F,warning=F,fig.width=14,fig.height=7}
par(mfrow = c(1,2))
plot(m1,which = 5)
plot(m2,which = 5)
```

---

# Regression Diagnostics

- Problematic points: points can be unusual, but not all unusual points are problems

--

- Consider three types of points:

--

  1. Outlier Point: an observation with a **large residual**
  
  2. Leverage Point: an observation with an extreme value for $x$
  
  3. Influential Point: an observation that changes the slope of the regression line
  
--

- Always good to look at **influential points** to ensure there isn't an error in the measurement

--

  - But NOT always necessary to blindly throw them out
  
  - Better to characterize how sensitive the results are to them

---

# Choosing Variables

- With all this in mind, **how do you choose your variables** and **specify your regression equation**?

--

- We know we want to specify the true relationships, but how do we do this in practice?

--

- Theory, *theory*, **theory** is essential and should come first

--

  - This can be formalized with models, or it can be described with intuition, but no amount of diagnostic plots can replace careful theorizing prior to analysis
  
--

- That being said, let's consider some additional tests

--

- One simple method is to compare two specifications, say one that includes $x_2$ as a control and another that doesn't

--

  - How can we compare these models?

---

# Goodness of Fit

- Recall the definition of the $R^2$ from the simple regression case

$$
\begin{aligned}
R^2 &= \frac{\sum(\hat{y}_i - \bar{y})^2}{\sum(y_i - \bar{y})^2}\\
&= \frac{SSE}{SST} \\
&= 1 - \frac{SSR}{SST}
\end{aligned}
$$
- where $SSE$ is the explained sum of squares and $SSR$ is the residual sum of squares

--

- The $R^2$ will never decrease as we add additional predictors

--

  - This is because the denominator doesn't change, but the numerator will either increase or stay the same with additional predictors
  
--

- This makes the $R^2$ a pretty terrible metric for comparing models!

---

# Goodness of Fit

- Instead, we typically use the adjusted $R$-square value:

$$
\begin{aligned}
R^2_{adj} &= 1 - \left[\frac{\frac{SSR}{(n-k-1)}}{\frac{SST}{(n-1)}}\right]\\
&= 1 - \frac{\frac{\hat{\sigma}^2_u}{(n-k-1)}}{\frac{SST}{(n-1)}}\\
\end{aligned}
$$
--

- By construction, this will only increase with a new predictor if that variable's $t$-statistic is greater than 1 in absolute value

--

$$
R^2_{adj} = 1 - \frac{(1-R^2)(n-1)}{(n-k-1)}
$$

--

- On your own, think about when the $R^2$ and $R^2_{adj}$ will be similar and different?

---

# Too many variables

- Define a variable (denoted $W$) as **irrelevant** if it has not partial effect on $y$ in the population: $\frac{\partial y}{\partial W} = 0$

--

- If we include $W$ in our model (aka "overspecifying the model") will not bias the estimates since it does not violate our assumptions 1 through 4

  - In other words, if the true model is $y = \beta_0 + \beta_1 x + u$ but we specify $y = \beta_0 + \beta_1 x + \beta_2 W + u$, all $\hat{\beta}$ will be unbiased
  
--

- However, we can still harm our model if $W$ is collinear with $x$

--

  - Recall that $Var(\hat{\beta}_j) = \frac{\sigma^2}{n*var(x_j)*(1 - R^2_j)}$ where $R^2_j$ is the $R$-squared obtained from regressing $x_j$ on all other independent variables in the model
  
  - If $W$ is correlated with $x$, then $Var(\hat{\beta}_1)$ will become inflated, meaning our model is less **efficient**
  
  - Put a different way, our statistical power decreases, increasing the likelihood of falsely accepting the null
  
---

# Too many variables

- You can assess this threat by calculating each $R^2_j$ yourself

--

  - Also can calculate the **variance inflation factor** (VIF): $VIF(\hat{\beta}_j) = \frac{1}{1 - R^2_j}$
  
  - Can rewrite $Var(\hat{\beta}_j) = \frac{\sigma^2}{n*var(x_j)}VIF(\hat{\beta}_j)$, which is where it gets it's name...the factor by which $Var(\hat{\beta}_j)$ is inflated due to the fact that $x_j$ is correlated with other $x$'s in the model
  
--

- However, we will sometimes *want* to include controls that are correlated with $y$ but are **not** correlated with $x$

--

  - Note that these are not necessary to recover unbiased estimates (remember the definition of OVB?)
  
--

- Why do we want to control for some $Z$ where $\frac{\partial x}{\partial Z} = 0$ but $\frac{\partial y}{\partial Z} \neq 0$?

--

  - It helps explain variation in $y$, meaning that $\sigma^2$ is lower, meaning $Var(\hat{\beta}_j)$ is also lower
  
  - In other words, it makes all our estimates more **efficient**

---

# Hypotheses about Parameters

- Thus far, we've always been implicitly interested in a single coefficient, or testing each one at a time

--

- But we might be interested in how two coefficients related to each other

  - For example, your book has the example where researchers are interested whether the effect on income of an additional year of education at a junior college is as much as the effect of an additional year of education at four-year university.
  
  - The idea here is that $jc$'s are lower status in the U.S. than universities, so maybe employers value these years of education less.
  
  - (A complementary hypothesis would be that a jc education may be of lower quality.) 
  
- The model assumed is $\log(wage) = \beta_{0}+\beta_{1}jc+\beta_{2}univ+\beta_{3}work+u$

---

# Hypotheses about Parameters

- If we are interested in whether there is a **difference** in returns to education from junior colleges and universities, what is the appropriate null hypothesis?

--

- $H_0: \beta_1 = \beta_2$

- And the alternative?

--

- $H_A: \beta_1 < \beta_2$

- We can re-write as:

$$
\begin{aligned}
H_0&: \beta_1 - \beta_2 = 0\\
H_A&: \beta_1 - \beta_2 < 0
\end{aligned}
$$

- Thus our quantity of interest is $\beta_1 - \beta_2$ and our test statistic is $\frac{\hat{\beta}_1 - \hat{\beta}_2}{se(\hat{\beta}_1 - \hat{\beta}_2)}$

---

# Hypotheses about Parameters

- What is $se(\hat{\beta}_1 - \hat{\beta}_2)$?

$$
\begin{aligned}
se(\hat{\beta}_1 - \hat{\beta}_2) &= \sqrt{var(\hat{\beta}_1 - \hat{\beta}_2)}~~~\text{and}\\
var(\hat{\beta}_1 - \hat{\beta}_2) &= var(\hat{\beta}_1) + var(\hat{\beta}_2) - 2cov(\hat{\beta}_1,\hat{\beta}_2) ~~\text{so}\\
se(\hat{\beta}_1 - \hat{\beta}_2) &= \sqrt{var(\hat{\beta}_1) + var(\hat{\beta}_2) - 2cov(\hat{\beta}_1,\hat{\beta}_2)}
\end{aligned}
$$
--

- We can just grab these values from the variance-covariance matrix of estimated betas

--

- Or we can do an even easier trick! Let's denote $\theta = \beta_1 - \beta_2$, meaning that $\beta_1 = \theta + \beta_2$

- Therefore:

$$
\begin{aligned}
\log (wage) &= \beta_{0}+\beta_{1}jc+\beta_{2}univ+\beta _{3}work+u\\
&= \beta_{0}+(\theta + \beta_2)jc+\beta_{2}univ+\beta _{3}work+u\\
&= \beta_{0}+\theta jc+\beta_{2}(univ+jc) + \beta _{3}work+u\\
\end{aligned}
$$
--

- So easy! Just create a new variable that is the sum of $univ$ and $jc$ and look at the coefficient on $jc$!

---

# Multiple Linear Restrictions


- What if we want to know if a **group** of predictors are **jointly** significant?

- Start by defining an **unrestricted** model as $UR: y = \beta_0 + \beta_1 x_1 + \dots + \beta_k x_k + u$

- Then take the group of variables we are interested in evaluating and move them to the end of the regression

--

  - I.e., if there are $q$ variables we want to test are jointly significant, denote these as $\beta_{k-q+1}, \beta_{k-q+2}$ etc.
  
  - Thus our null hypothesis is $H_0: \beta_{k-q+1} = \beta_{k-q+2} = \dots = \beta_k = 0$
  
--

- We can write a restricted model as $R: y = \beta_0 + \beta_1 x_1 + \dots + \beta_{k-q}x_{k-q} + u$

- To test, we use the $F$-statistic defined as $F \equiv \frac{(SSR_r - SSR_{ur})/q}{SSR_{ur}/ (n-k-1)}$

--

  - Note that, since $SSR_r \geq SSR_{ur}$, $F > 0$
  
--

- This is the ratio of two independent $\chi^2$ random variables, divided by their respective degrees of freedom

- We can therefore conduct hypothesis testing using this: if it is extremely unlikely that we would obtain the observed $F$-statistic by chance, we reject the null $H_0$

```{r,message=F,echo=F,warning=F,results='hide'}
dir <- getwd()
type <- 'pdf'
format <- 'landscape'
f <- 'Lecture_20_slides'

system(paste('Rscript ../NFGH/chromeprint.R',dir,type,format,f),wait = F)
```
