---
title: "Lecture 13"
subtitle: "Quantitative Political Science"
author: "Prof. Bisbee"
institute: "Vanderbilt University"
date: "Lecture Date: 2023/10/24\n Slides Updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    # self_contained: true
    chakra: libs/remark-latest.min.js
    lib_dir: libs
    css:
      - default
      - css/lexis.css
      - css/lexis-fonts.css
    #seal: false
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "17:9"

---

```{css,echo = F}
.small .remark-code { /*Change made here*/
  font-size: 85% !important;
}
.tiny .remark-code { /*Change made here*/
  font-size: 50% !important;
}
```

```{r,include=F}
set.seed(123)
options(width=60)
knitr::opts_chunk$set(fig.align='center',fig.width=9,fig.height=5)
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
require(tidyverse)
require(ggpubr)
```

# Agenda

1. Associations

2. Describing relationships

3. Modeling relationships


---
  
# Associations between two (or more) variables
  
- Thus far, description & inference for either one variable or two variables drawn from different units

- Now, what you are here for! Relationships between two variables from the *same* units!
  
- Bivariate:
  
  - $X$ (independent / predictor / explanatory / RHS) and $Y$ (dependent / outcome)

  - Implies a causal intuition, but does NOT buy it!
  
- 4 approaches (more like steps)

  1. Displaying relationships
  
  2. Summarizing **non-parametrically**
  
  3. Summarizing **parametrically**
  
  4. Making **inferences** about the relationship in a population from a sample
  
---

# 1. Displaying Relationships

- Crosstab(ulation)s: $Y$ typically in rows, $X$ in columns

```{r}
require(tidyverse)
dat <- read_rds('https://github.com/jbisbee1/PSCI_8356/raw/main/Lectures/Data/sc_debt.Rds')

t <- table(dat$preddeg,dat$control)
t
```

---

# 1. Displaying Relationships

- Crosstab(ulation)s: $Y$ typically in rows, $X$ in columns

```{r}
prop.table(t,margin = 1) # Rows
prop.table(t,margin = 2) # Columns
```

---

# 1. Displaying Relationships

- Crosstab(ulation)s: $Y$ typically in rows, $X$ in columns...BAD FOR MANY CATEGORIES OR CONTINUOUS!

```{r}
table(dat$md_earn_wne_p6,dat$sat_avg)
```

---

# 1. Displaying Relationships

- Use bins first

```{r}
dat <- dat %>%
  mutate(earn_quartiles = cut(md_earn_wne_p6,breaks = quantile(md_earn_wne_p6,p = c(0,.25,.5,.75,1),na.rm=T),dig.lab = 10),
         sat_quartiles = cut(sat_avg,breaks = quantile(sat_avg,p = c(0,.25,.5,.75,1),na.rm=T),dig.lab = 10))
t <- table(dat$earn_quartiles,dat$sat_quartiles)
round(prop.table(t,margin = 2)*100,2)
```

---

# 1. Displaying Relationships

- Plotting

  - Barplots (`geom_bar()`): $X$ and $Y$ are both categorical (including binary)
  
  - Densities / histograms (`geom_density()` / `geom_histogram()`): $X$ is binary and $Y$ is continuous
  
  - Boxplots / violin plots (`geom_boxplot()` / `geom_violin()`): $X$ is categorical and $Y$ is continuous
  
  - Scatterplots (`geom_point()`): $X$ and $Y$ are both continuous

---

# 1. Displaying Relationships

  
- Barplots (`geom_bar()`): $X$ and $Y$ are both categorical (including binary)

```{r,message=F,warning=F}
dat %>%
  ggplot(aes(x = control,fill = preddeg)) + 
  geom_bar()
```

---

# 1. Displaying Relationships

  
- Densities / histograms (`geom_density()` / `geom_histogram()`): $X$ is binary and $Y$ is continuous

```{r,message=F,warning=F}
dat %>%
  ggplot(aes(x = sat_avg,color = control)) + 
  geom_density()
```

---

# 1. Displaying Relationships

  
- Densities / histograms (`geom_density()` / `geom_histogram()`): $X$ is binary and $Y$ is continuous

```{r,message=F,warning=F}
dat %>%
  ggplot(aes(x = sat_avg,fill = region)) + 
  geom_histogram(position = 'dodge')
```

---

# 1. Displaying Relationships

  
- Boxplots / violin plots (`geom_boxplot()` / `geom_violin()`): $X$ is categorical and $Y$ is continuous

```{r,message=F,warning=F}
helper <- dat %>%
  group_by(region) %>%
  summarise(medSAT = median(sat_avg,na.rm=T))
dat %>%
  mutate(region = factor(region,levels = helper %>% arrange(medSAT) %>% pull(region))) %>%
  ggplot(aes(x = sat_avg,y = reorder(region,sat_avg))) + 
  geom_boxplot(outlier.colour = 'red')


```

---

# 1. Displaying Relationships

  
- Boxplots / violin plots (`geom_boxplot()` / `geom_violin()`): $X$ is categorical and $Y$ is continuous

```{r,message=F,warning=F}
dat %>%
  ggplot(aes(x = sat_avg,y = reorder(region,sat_avg))) + 
  geom_violin()
```

---

# 1. Displaying Relationships

  
- Scatterplots (`geom_point()`): $X$ and $Y$ are both continuous

```{r,message=F,warning=F}
dat %>%
  filter(md_earn_wne_p6 > 70000,control == 'Public')
dat %>%
  ggplot(aes(x = costt4_a,y = md_earn_wne_p6,color = control)) + 
  geom_point() + 
  geom_smooth(method = 'lm')
```

---

# 2. Summarizing Non-parametrically

- Conditional means

  - What is the average value of $Y$ for a given value of $X$?
  
```{r}
dat %>%
  drop_na(sat_avg) %>%
  # mutate()
  group_by(sat_avg) %>%
  summarise(mean_earn = mean(md_earn_wne_p6,na.rm=T))
```

---

# 2. Summarizing Non-parametrically

- Conditional means

  - What is the average value of $Y$ for a given value of $X$?
  
```{r,message=F,warning=F}
p <- dat %>%
  drop_na(sat_quartiles) %>%
  group_by(sat_quartiles) %>%
  summarise(mean_earn = mean(md_earn_wne_p6,na.rm=T)) %>%
  ggplot(aes(x = sat_quartiles,y = mean_earn)) + 
  geom_bar(stat = 'identity')
```

---

# 2. Summarizing Non-parametrically

- Conditional means

  - What is the average value of $Y$ for a given value of $X$?
  
```{r,message=F,warning=F}
p
```

---

# 2. Summarizing Non-parametrically

- Conditional means

  - What is the average value of $Y$ for a given value of $X$?
  
```{r,message=F,warning=F}
p + geom_text(aes(label = scales::dollar(round(mean_earn))),vjust = -.5)
```

---

# 2. Summarizing Non-parametrically

- Smoothers
  
```{r,message=F,warning=F}
dat %>%
  ggplot(aes(x = sat_avg,y = md_earn_wne_p6)) + 
  geom_point() + geom_smooth()
```

---

# 3. Summarizing Parametrically

- Want to use *models* to describe **theoretical** relationships

- Want minimal **assumptions**...thus far?

- For inferences about $\mu$ with **large** samples

  - **identicality**: necessary for $\bar{Y}$ to be unbiased for $\mu$
  
  - **independence**: necessary for $VAR(\bar{Y}) = \frac{\sigma^2}{n}$
  
- For inferences about $\mu$ with **small** samples

  - $Y \sim \mathcal{N}(\mu,\sigma^2)$
  
---

# 3. Summarizing Parametrically

- For inferences about differences in population means with **large** samples

  - Two samples are drawn **independently**
  
- For inferences about differences in population means with **large** samples

  - Two samples are drawn **independently**
  
  - Two samples have the **same variance**
  
  - Underlying populations are **Normal**
  
- This is a pretty short list!

- Lots more to come with bivariate and multivariate analysis!

---

# 3. Summarizing Parametrically

- How to describe a bivariate relationship?

- Start with notion of **correlation**

$$
\begin{aligned}
\rho &= \frac{COV(Y_1,Y_2)}{\sigma_1\sigma_2} \\
&= \frac{E[(Y_1 - \mu_1)(Y_2 - \mu_2)]}{\sigma_1\sigma_2}
\end{aligned}
$$

- Translating to bivariate world is easy, just use $X$ and $Y$

$$
\rho = \frac{E[(X - \mu_X)(Y - \mu_Y)]}{\sigma_X\sigma_Y}
$$

- But $\rho$ is a theoretical quantity (a **parameter**)

- What is a good estimator?

---

# 3. Summarizing Parametrically

$$
r = \frac{\sum_i(X_i - \bar{X})(Y_i - \bar{Y})}{\sqrt{\sum_i (X_i - \bar{X})^2 \sum_i(Y_i - \bar{Y})^2}}
$$
- Replace the covariance with the sample covariance $s_{XY} = \sum_i(X_i - \bar{X})(Y_i - \bar{Y})$ and the standard deviations for both variables with their sample analogues $s_X = \sqrt{\sum_i (X_i - \bar{X})^2}$ and $s_Y = \sqrt{\sum_i (Y_i - \bar{Y})^2}$.

- How good is this? It depends on the underlying data

```{r}
X <- seq(-100,100,by = 1)
```


---

# 3. Summarizing Parametrically

- Works well with linear relationships

```{r}
Y <- X + rnorm(length(X),mean = 0,sd = 5)
data.frame(X = X,Y = Y) %>%
  ggplot(aes(x = X,y = Y)) + geom_point() + stat_cor(p.digits = NA,label.sep = '')
```

---

# 3. Summarizing Parametrically

- Still works ok with more noise

```{r}
Y <- X + rnorm(length(X),mean = 0,sd = 50)
data.frame(X = X,Y = Y) %>%
  ggplot(aes(x = X,y = Y)) + geom_point() + stat_cor(p.digits = NA,label.sep = '')
```

---

# 3. Summarizing Parametrically

- Doesn't work well with curvelinear relationships

```{r}
Y <- X^2 + rnorm(length(X),mean = 0,sd = 1000)
data.frame(X = X,Y = Y) %>%
  ggplot(aes(x = X,y = Y)) + geom_point() + stat_cor(p.digits = NA,label.sep = '')
```

---

# 3. Summarizing Parametrically

- Very sensitive to outliers

```{r}
X <- rnorm(100,mean = 33,sd = 3)
Y <- rnorm(100,mean = 50,sd = 3)
data.frame(X = X,Y = Y) %>%
  ggplot(aes(x = X,y = Y)) + geom_point() + stat_cor(p.digits = NA,label.sep = '') + lims(x = c(0,100),y = c(0,100))
```

---

# 3. Summarizing Parametrically

- Very sensitive to outliers

```{r}
X[75] <- Y[75] <- 75
data.frame(X = X,Y = Y) %>%
  ggplot(aes(x = X,y = Y)) + geom_point() + stat_cor(p.digits = NA,label.sep = '') + 
  xlim(c(0,100)) + ylim(c(0,100))
```


```{r,message=F,echo=F,warning=F,results='hide'}
dir <- getwd()
type <- 'pdf'
format <- 'landscape'
f <- 'Lecture_13_slides'

system(paste('Rscript ../NFGH/chromeprint.R',dir,type,format,f),wait = F)
```
