---
title: "Lecture Wrapup"
subtitle: "Quantitative Political Science"
author: "Prof. Bisbee"
institute: "Vanderbilt University"
date: "Lecture Date: 2024/12/03\n Slides Updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    # self_contained: true
    chakra: libs/remark-latest.min.js
    lib_dir: libs
    css:
      - default
      - css/lexis.css
      - css/lexis-fonts.css
    #seal: false
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "17:9"

---

```{css,echo = F}
.small .remark-code { /*Change made here*/
  font-size: 85% !important;
}
.tiny .remark-code { /*Change made here*/
  font-size: 50% !important;
}
```

```{r,include=F}
set.seed(123)
options(width=60)
knitr::opts_chunk$set(fig.align='center',fig.width=9,fig.height=5)
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
require(tidyverse)
require(ggpubr)
```

# Agenda

1. Variance of OLS estimators

2. Heteroskedasticity

---

# Recap

- We went over 4 assumptions to characterize the bias of our OLS estimators

1. Relationship between $x$ and $y$ is **linear in its parameters**

2. $x$ and $y$ are drawn from a random sample, making them **i.i.d.**

3. $VAR(X) \neq 0$

4. $E(u|x) = 0$

- With these, we demonstrated that $\hat{\beta}_0$ and $\hat{\beta}_1$ are **unbiased** for $\beta_0$ and $\beta_1$

---

# Sampling Distributions

- $\hat{\beta}_0$ and $\hat{\beta}_1$ are statistics, just like $\bar{Y}$

- When we evaluate statistics, we care about both their **bias** (last class) and their **variance**

  - How far can we expect them to be from their true value (i.e., the population parameter) on average?
  
- In the univariate case, we were interested in the **sampling distribution** of $\bar{Y}$

- Here, we are also interested in the sampling distributions of $\hat{\beta}_0$ and $\hat{\beta}_1$

---

# Variance

- We already know the means of $\hat{\beta}_0$ and $\hat{\beta}_1$: they are $\beta_0$ and $\beta_1$ (from last class)

- To compute the **variances** of $\hat{\beta}_0$ and $\hat{\beta}_1$, we need a **fifth assumption**

Assumption 5: $VAR(u|x) = \sigma^2$

  - The error has the **same variance** regardless of the value of $x$
  
- This is known as **homoskedasticity**

- If this fails, we have **heteroskedasticity**

---

# Variance

```{r,message=F,warning =F}
require(tidyverse)
set.seed(123)
X <- rnorm(500,mean = 12,sd = 4)
Y <- rnorm(500,mean = X,sd = X/3)
Y2 <- rnorm(500,mean = X,sd = 2)

p <- data.frame(X = X,Heteroskedastic = Y,Homoskedastic = Y2) %>%
  gather(outcome,value,-X) %>%
  mutate(outcome = factor(outcome,levels = c('Homoskedastic','Heteroskedastic'))) %>%
  ggplot(aes(x = X,y = value)) + 
  geom_point() + 
  geom_smooth(method = 'lm') + 
  facet_wrap(~outcome)
```

---

# Variance

```{r,message=F,warning =F}
p
```

---

# Variance

- Note the difference between Assumptions 4 and 5!

  - $E(u|x) = 0$
  
  - $VAR(u|x) = \sigma^2$
  
- We don't need assumption 5 for unbiasedness, but we do for variance!

- What is $VAR(y|x)$?

---

# Variance

$$
\begin{aligned}
VAR(y|x) &= VAR(\beta_0 + \beta_1 x + u | x)\\
&= VAR(\beta_0 | x) + VAR(\beta_1 x | x) + VAR(u | x)\\
&= 0 + 0 + \sigma^2\\
&= \sigma^2
\end{aligned}
$$
- What is $\sigma^2$?

--

- It is a measure of the **extent to which unexplained factors are affecting $y$**

  - These factors are not related to $x$ (from assumption 4)
  
  - These factors are constant regardless of $x$ (from assumption 5)
  
  - When $\sigma^2$ is big, it means that other factors explain a lot of variation in $y$ beyond just $x$
  
  - When $\sigma^2$ is small, it means that $x$ explains a lot of variation in $y$
  
- Note that $\sigma^2$ is a **parameter**, something that exists in the population
  
---

# Variance of estimators

- $VAR(\hat{\beta}_0) = \frac{\sigma^2 \frac{\sum x_i^2}{n}}{SST_x}$ and $VAR(\hat{\beta}_1) = \frac{\sigma^2}{SST_x}$

- I'll leave it to you to prove $VAR(\hat{\beta}_0)$, but let's dig into $VAR(\hat{\beta}_1)$

$$
\begin{aligned}
\hat{\beta}_1 &= \beta_1 + \frac{\sum(x_i - \bar{x})u_i}{SST_x} \\
VAR(\hat{\beta}_1|x) &= VAR\bigg[\beta_1 + \frac{\sum(x_i - \bar{x})u_i}{SST_x}~\bigg|~x\bigg]\\
&= VAR(\beta_1|x) + \frac{1}{SST_x^2}\sum(x_i-\bar{x})^2VAR(u_i~|~x)\\
&= 0 + \frac{SST_x}{SST_x^2}VAR(u_i|x)\\
&= \frac{\sigma^2}{SST_x}
\end{aligned}
$$

---

# Sampling Variance of $\hat{\beta}_1$

- So $VAR(\hat{\beta}_1) = \frac{\sigma^2}{SST_x}$

- Want this to be as small as possible (bias-variance tradeoff)

- As $\sigma^2$ gets smaller, so does $VAR(\hat{\beta}_1)$

- As $SST_x$ gets bigger, $VAR(\hat{\beta}_1)$ gets smaller

- Unpack $SST_x$ for more insights!

---

# Sampling Variance of $\hat{\beta}_1$

$$
\begin{aligned}
SST_x &= \sum(x_i - \bar{x})^2\\
\frac{SST_x}{n} &= \frac{\sum(x_i - \bar{x})^2}{n}\\
&= var(x) \text{ sample variance of }x\\
SST_x &= n*var\\
VAR(\hat{\beta}_1) &= \frac{\sigma^2}{n*var(x)}
\end{aligned}
$$
- What can we actually manipulate here as a researcher?

  - $\sigma^2$ is a parameter: it declines when $x$ explains $y$ well. But we don't have a lot of control over this.
  
  - $var(x)$ is the empirical variance of $x$ in our sample. It approximates the population variance, but we don't have a ton of control over this either.
  
  - $n$: we choose this!
  
---

# Heteroskedasticity

- The preceding results rely on the assumption of $VAR(u|x) = \sigma^2$

- What if this doesn't hold?

```{r,message=F,warning =F}
p
```

---

# Heteroskedasticity

- We would then say that the variance of the errors conditional on $x$ is specific to that unit

  - $VAR(u_i|x_i) = \sigma_i^2$
  
- Recall that $\hat{\beta}_1 = \frac{\sum(x_i - \bar{x})u_i}{\sum(x_i - \bar{x})^2}$ and that 

$$
\begin{aligned}
VAR(\hat{\beta}_1) &= VAR\bigg[\frac{\sum(x_i - \bar{x})u_i}{\sum(x_i - \bar{x})^2}\bigg]\\
&= \frac{1}{SST_x^2}\sum(x_i - \bar{x})^2 VAR(u_i|x_i)\\
&= \frac{\sum(x_i - \bar{x})^2\sigma_i^2}{SST_x^2}
\end{aligned}
$$


---

# Heteroskedasticity

- What to do?

- In 1980, one of the most cited economics papers was written by Halbert White

- In it, he proposed calculating heteroskedastic robust standard errors as $\widehat{VAR(\hat{\beta}_1)} = \frac{\sum(x_i - \bar{x})^2\hat{u}_i^2}{SST_x^2}$

  - $\hat{u}_i^2$ is just the squared residual associated with each observation $i$
  
- These standard errors have LOTS of different names:

  - "White standard errors"
  
  - "Huber-White standard errors"
  
  - "Robust standard errors"
  
  - "Heteroskedasticity-robust standard errors"
  
---

# Stepping Back

- The easiest thing to do is just calculate robust standard errors with $\widehat{VAR(\hat{\beta}_1)} = \frac{\sum(x_i - \bar{x})^2\hat{u}_i^2}{SST_x^2}$

- However, if our fifth assumption fails (remember: $VAR(u|x) = \sigma^2$), it means that the OLS estimator is no longer the ``best'' estimator

- What do we mean by ``best''? The lowest-variance!

  - As an aside, **B**est **L**inear **U**nbiased **E**stimator or BLUE is a commonly used acronym for the OLS estimators. Note that we can prove **U**nbiasedness with the first four assumptions, and assumption 5 gives us **B**est
  
---

# Estimating Error Variance

- So we have $VAR(\hat{\beta}_1) = \frac{\sigma^2}{SST_x}$

  - This is a **conceptual** quantity
  
- How do we actually calculate it?

  - Recall from the univariate case where we wrote $VAR(\bar{Y}) = \frac{\sigma^2}{n}$
  
  - We said it is rare that we actually know $\sigma^2$, but we still estimate it with $S_u^2 = \frac{\sum(y_i - \bar{y})^2}{n-1}$
  
- Here, we do something very similar: $\hat{\sigma}^2 = \frac{\sum(u_i)^2}{n-2} = \frac{SSR}{n-2}$

  - Why $n-2$?
  
---

# Estimating Error Variance

- Thus we plug this into our two formulas for $VAR(\hat{\beta}_0)$ and $VAR(\hat{\beta}_1)$

$$
\begin{aligned}
\widehat{VAR(\hat{\beta}_0)} &= \frac{\hat{\sigma}^2 \frac{\sum x_i^2}{n}}{SST_x}\\
&= \frac{\frac{SSR}{n-2}\frac{\sum x_i^2}{n}}{SST_x}\\
\widehat{VAR(\hat{\beta}_1)} &= \frac{\hat{\sigma}^2}{SST_x}\\
&= \frac{\frac{SSR}{n-2}}{SST_x}\\
\end{aligned}
$$

---

# Estimating Error Variance

- As we discussed last week in the theoretical case, $\hat{\sigma}^2$ is a very interesting quantity, because $\sqrt{\hat{\sigma}^2} = \hat{\sigma} \overset{p}\rightarrow \sigma$

  - $\hat{\sigma}$ is expressed in units of $y$
  
  - Tell us how far the typical fitted value of $y$ is from the observed value
  
  - Theoretically, the extent to which unexplained factors are affecting the value of $y$
  
  - **VERY INFORMATIVE STATISTIC THAT NO ONE REALLY PAYS ATTENTION TO**
  
- Terms for $\hat{\sigma}$:

  - Wooldridge: "standard error of regression" (SER)
  
  - Root MSE or RMSE
  
  - Standard error of the estimate (SEE)
  
  - `R`: Residual standard error



---

# Hypothesis Testing

- Remember all these fun times we had?

```{r,echo=F,warning=F,message=F}
# Load necessary library
library(ggplot2)

# Define the mean and standard deviation
mean <- 0
sd <- 1

# Define the range for x values
x <- seq(-4, 4, length.out = 1000)

# Define the density function for the normal distribution
y <- dnorm(x, mean, sd)

# Calculate the critical value for a 95% confidence interval
z <- qnorm(c(0.025, 0.975))

ggplot(data.frame(x, y), aes(x, y)) +
  geom_ribbon(data = subset(data.frame(x, y), x > -Inf & x < Inf),
              aes(ymin = 0, ymax = y),
              fill = "gray40",
              alpha = 0.5,color = 'black') +
  geom_ribbon(data = subset(data.frame(x, y), x > z[1] & x < z[2]),
              aes(ymin = 0, ymax = y),
              fill = "white",
              alpha = 1,color = 'black') +
  labs(title = bquote("Sampling Distribution of"~ hat(theta)),
       x = bquote(z),
       y = bquote(f(hat(theta))==phi)) +
  theme_minimal() + 
  geom_vline(xintercept = 0,linetype = 'dashed') + 
  annotate(geom = 'segment',x = 1.96,y = -.05,xend = Inf,yend = -.05,
             size = 1.6,color = 'gray40') +
  annotate(geom = 'segment',x = -1.96,y = -.05,xend = -Inf,yend = -.05,
             size = 1.6,color = 'gray40') +
  annotate(geom = 'segment',x = -Inf,y = -.05,xend = Inf,yend = -.05,
           arrow = arrow(length=unit(0.30,"cm"), ends="both", type = "closed")) +
  annotate(geom = 'segment',x = 1.96,y = -.075,xend = 1.96,yend = -0.025) + 
  annotate(geom = 'segment',x = -1.96,y = -.075,xend = -1.96,yend = -0.025) + 
  annotate(geom = 'label',x = 0,y = -Inf,label = expression(theta[0]),vjust = 0,parse = T) + 
  # annotate(geom = 'segment',x = -.95,xend = .95,y = .25,yend= .25,
  #          arrow = arrow(length=unit(0.30,"cm"), ends="both", type = "closed")) + 
  annotate(geom = 'label',x = 0,y = Inf,label = expression(1-alpha),vjust = 5.5,parse = T) + 
  annotate(geom = 'label',x = -2.5,y = .01,label = expression(alpha/2),vjust = 0,parse = T) + 
  annotate(geom = 'label',x = 2.5,y = .01,label = expression(alpha/2),vjust = 0,parse = T) + 
  scale_x_continuous(breaks = c(-1.96,1.96),labels = c(expression(theta[0]-z[alpha/2]~sigma[hat(theta)]),expression(theta[0]+z[alpha/2]~sigma[hat(theta)]))) + 
  theme(axis.title.x = element_text(hjust = 1))
```

---

# Hypothesis Testing

- We now have the tools to do this with $\hat{\beta}_1$! (And $\hat{\beta}_0$, although that is rarely the quantity of interest.)

- Note that we typically are interested in whether $\hat{\beta}_1$ is zero:

  - Null $H_0$: $\beta_1 = 0$
  
  - Alternative $H_A$: $\beta_1 \neq 0$
  
  - Test statistic: Critical $t$ value for Student's T-test for our estimator $\hat{\beta}_1$
  
  - Rejection Region: $\hat{\beta}_1 < 0 - t_{\alpha/2,\nu}*\sqrt{\widehat{VAR(\hat{\beta}_1)}}$ or $\hat{\beta}_1 > 0 + t_{\alpha/2,\nu}*\sqrt{\widehat{VAR(\hat{\beta}_1)}}$
  
---

# Hypothesis Testing

- What is $\sqrt{\widehat{VAR(\hat{\beta}_1)}}$? 

--

  - The **standard error** of the estimator $\hat{\beta}_1$, or $\widehat{se(\hat{\beta}_1)}$, (or often just $se_{\hat{\beta}_1}$)
  
- What is $\nu$?

--

  - The **degrees of freedom**: This will be $n - k - 1$. $n$ observations minus $k$ parameters (in this case just one: $\hat{\beta}_1$) - 1 (for the intercept $\hat{\beta}_0$)

---

# Hypothesis Testing


```{r,echo=F,warning=F,message=F}
# Load necessary library
library(ggplot2)

# Define the mean and standard deviation
mean <- 0
sd <- 1

# Define the range for x values
x <- seq(-4, 4, length.out = 1000)

# Define the density function for the normal distribution
y <- dnorm(x, mean, sd)

# Calculate the critical value for a 95% confidence interval
z <- qnorm(c(0.025, 0.975))

ggplot(data.frame(x, y), aes(x, y)) +
  geom_ribbon(data = subset(data.frame(x, y), x > -Inf & x < Inf),
              aes(ymin = 0, ymax = y),
              fill = "gray40",
              alpha = 0.5,color = 'black') +
  geom_ribbon(data = subset(data.frame(x, y), x > z[1] & x < z[2]),
              aes(ymin = 0, ymax = y),
              fill = "white",
              alpha = 1,color = 'black') +
  labs(title = bquote("Sampling Distribution of"~ hat(beta)[1]),
       x = bquote(t),
       y = bquote(f(hat(theta))==phi)) +
  theme_minimal() + 
  geom_vline(xintercept = 0,linetype = 'dashed') + 
  annotate(geom = 'segment',x = 1.96,y = -.05,xend = Inf,yend = -.05,
             size = 1.6,color = 'gray40') +
  annotate(geom = 'segment',x = -1.96,y = -.05,xend = -Inf,yend = -.05,
             size = 1.6,color = 'gray40') +
  annotate(geom = 'segment',x = -Inf,y = -.05,xend = Inf,yend = -.05,
           arrow = arrow(length=unit(0.30,"cm"), ends="both", type = "closed")) +
  annotate(geom = 'segment',x = 1.96,y = -.075,xend = 1.96,yend = -0.025) + 
  annotate(geom = 'segment',x = -1.96,y = -.075,xend = -1.96,yend = -0.025) + 
  annotate(geom = 'label',x = 0,y = -Inf,label = expression(beta[1]==0),vjust = 0,parse = T) + 
  # annotate(geom = 'segment',x = -.95,xend = .95,y = .25,yend= .25,
  #          arrow = arrow(length=unit(0.30,"cm"), ends="both", type = "closed")) + 
  annotate(geom = 'label',x = 0,y = Inf,label = expression(1-alpha),vjust = 5.5,parse = T) + 
  annotate(geom = 'label',x = -2.5,y = .01,label = expression(alpha/2),vjust = 0,parse = T) + 
  annotate(geom = 'label',x = 2.5,y = .01,label = expression(alpha/2),vjust = 0,parse = T) + 
  scale_x_continuous(breaks = c(-1.96,1.96),labels = c(expression(0-t[alpha/2~nu]~se[hat(beta)[1]]),expression(0+t[alpha/2~nu]~se[hat(beta)[1]]))) + 
  theme(axis.title.x = element_text(hjust = 1))
```

---

```{r}
require(tidyverse)
set.seed(123)
n <- 100
X <- rnorm(n)
Y <- rnorm(n,mean = X)

summary(lm(Y~X))
```

---

# Manual Calculation!

```{r}
b1_hat <- cov(X,Y)/var(X)
b0_hat <- mean(Y) - (cov(X,Y)/var(X))*mean(X)

preds <- b0_hat + b1_hat*X
resids <- Y - preds
mean(resids)
```

---

# Manual Calculation!

```{r}
SSR <- sum(resids^2)
sigma2_hat <- SSR/(n-2)
sigma_hat <- sqrt(sigma2_hat)

SST_x <- sum((X - mean(X))^2)
S_xx <- sum(X^2) - n*mean(X)^2 # Equivalent ways

VAR0_hat <- (sigma2_hat*(sum(X^2)/n))/SST_x
se0_hat <- sqrt(VAR0_hat)

VAR1_hat <- sigma2_hat/SST_x
se1_hat <- sqrt(VAR1_hat)
```

---

# Manual Calculation!

```{r}
cat(c(b0_hat,se0_hat),'\n',c(b1_hat,se1_hat))

summary(lm(Y~X))
```

---

# Other output

- Regression output includes two additional columns

  - `t value` and `Pr(>|t|)`
  
- `t value` is just $\frac{\hat{\beta}_1}{\widehat{se(\hat{\beta}_1)}}$ (or $\frac{\hat{\beta}_0}{\widehat{se(\hat{\beta}_0)}}$)

  - `b1_hat / se1_hat = ` `r b1_hat / se1_hat`
  
- `Pr(>|t|)` is literally the probably of observing a value as large as the absolute value of the t-value

  - I.e., the $p$**-value**!: "attained significance level" or "smallest level of $\alpha$ for which we would **reject** $H_0$

---

# Controlling for a variable

- We talked last week about OVB (**O**mitted **V**ariable **B**ias)

  - Subset of broader conversation about bias
  
- We might want to "control" for $z$ to remove OVB (preventing the $\beta_2 z_i + \nu_i$ from being "buried in the $u$)

  - But remember that failing to control for $z$ is only a problem **if** either (1) $\beta_2 \neq 0$ or (2) $cov(z,x) \neq 0$
  
- *Ceteris paribus*: "all things being equal"
  
  - Want to estimate a *ceteris paribus* relationship between $X$ and $Y$
  
  - What would the relationship look like if all other aspects of our units were the same?
  
  - Commonly invoked for causal claims, but more on that next semester
  
---

# Controlling for a variable

- Let's get precise with our terminology:

  - $Z$ is a potential **confound**
  
  - If $Z$ "confounds" the relationship between $X$ and $Y$, it **renders the relationship spurious**

- How about some examples? 

--

| $X$ | $Y$ | $Z$ | 
| ------------ | ------------------- | -------------------- |
| College degree | Salary at age 25 | Ability |
| Female | Pro-Choice | Democrat |
| First-born | IQ Score | Parental involvement | 
| Asian-American | Trump Support | Vietnamese | 
| Own a home | Participated in Women's March | Year of Birth |
| ... | ... | ... |

---

# Controlling for a variable

- To determine whether $Z$ renders the relationship between $X$ and $Y$ spurious, we...

  - "control for $Z$"
  
  - "condition on $Z$"
  
  - "hold $Z$ constant"
  
- These all mean the same thing conceptually, but there are several different ways to do this

- Ideally, we would do exactly what "holding $Z$ constant" suggests: divide our units by categories of $Z$ and examine the relationship between $X$ and $Y$ within each category of $Z$

  - I.e., if women are more pro-choice, we want to see this among *both* Democrats and Republicans
  
  - If the relationship persists after holding $Z$ constant, we say it is not spurious
  
  - If it no longer holds, we say that $Z$ is a confound rendering the relationship between $X$ and $Y$ spurious
  
- In practice, we usually do something much less careful

---

# Controlling for a variable

- We often will make our assumptions explicit with a **D**irected **A**cyclic **G**raph (DAG)

  - This encodes our intuition about what the population parameters of interest might be

```{r test,echo=FALSE,message=F,warning=F}
# library(DiagrammeR)
# m <- mermaid("graph LR;
# 		   X-->Y;
# 		   Z-->X;
# 		   Z-->Y
# 		   ")
# m
# widgetframe::frameWidget(m)

require(ggdag)
simple_dag_with_coords <- dagify(
  Y ~ X + Z,
  X ~ Z,
  # exposure = "x",
  # outcome = "y",
  coords = list(x = c(X = 1, Z = 2, Y = 3),
                y = c(X = 2, Z = 3, Y = 2))
)

ggdag(simple_dag_with_coords,text_size = 5) +
  theme_dag()
```

---

# Controlling for a variable

- Let's tackle a classic: education and income

  - $Y$: income
  
  - $X$: education
  
- What is $Z$?

--

  - Parent's education?

---

# Controlling for a variable

- Start by looking at all three relationships separately

```{r,message=F,warning=F}
require(tidyverse)
require(haven) # To open .dta files

dat <- read_dta('../NFGH/GSS_stata/GSS_stata/gss7222_r1.dta')

dat <- dat %>%
  select(realinc,educ,paeduc) %>%
  sample_n(size = 10000,replace = F) %>%
  drop_na()

gc()
```

- (`gc()` helps save memory after I dropped thousands of rows and columns)

---

# Controlling for a variable

- Start by looking at all three relationships separately

```{r,message=F,warning=F}
dat %>%
  group_by(educ) %>%
  summarise(income = mean(realinc,na.rm=T))
```

---

# Controlling for a variable

- Start by looking at all three relationships separately

```{r,message=F,warning=F}
dat %>%
  ggplot(aes(x = educ,y = realinc)) + 
  geom_point()
```

---

# Controlling for a variable

- Start by looking at all three relationships separately

```{r,message=F,warning=F,echo=F}
dat %>%
  ggplot(aes(x = educ,y = realinc)) + 
  geom_jitter() + 
  geom_smooth(aes(color = 'Local linear smoother')) + 
  geom_smooth(aes(color = 'Linear regression'),method = 'lm',linetype = 'dashed') + 
  scale_color_manual(name = 'Method',values = c('darkblue','red'))
```

---

# Controlling for a variable

- Start by looking at all three relationships separately

```{r,message=F,warning=F,echo=F}
dat %>%
  ggplot(aes(x = paeduc,y = realinc)) + 
  geom_jitter() + 
  geom_smooth(aes(color = 'Local linear smoother')) + 
  geom_smooth(aes(color = 'Linear regression'),method = 'lm',linetype = 'dashed') + 
  scale_color_manual(name = 'Method',values = c('darkblue','red'))
```

---

# Controlling for a variable

- Start by looking at all three relationships separately

```{r,message=F,warning=F,echo=F}
dat %>%
  ggplot(aes(x = paeduc,y = educ)) + 
  geom_jitter() + 
  geom_smooth(aes(color = 'Local linear smoother')) + 
  geom_smooth(aes(color = 'Linear regression'),method = 'lm',linetype = 'dashed') + 
  scale_color_manual(name = 'Method',values = c('darkblue','red'))
```

---

# Controlling for a variable

- Clearly evidence of:

  - $\beta_1 \neq 0$
  
  - $\beta_2 \neq 0$
  
  - $cov(educ,paeduc) \neq 0$
  
- I.e., OVB!

- Think through what this will mean for the following regression: $realinc_{i} = \beta_0 + \beta_1 educ_{i} + u_{i}$

---

# Controlling for a variable

- Let's "control" for parent's education in three ways

1. Fewest Assumptions: Just visualize it with a local linear smoother, and subset to different values of parent's education

  - No linearity assumption between $X$ and $Y$
  
  - Different relationship between $X$ and $Y$ for different values of $Z$

2. More Assumptions: Run multiple linear regressions, subsetting to different values of parent's education

  - Linearity assumption between $X$ and $Y$
  
  - Different (linear) relationships between $X$ and $Y$ for different values of $Z$
  
3. Most Assumptions: Run single linear regression, adding $Z$ as an additional predictor

  - Linearity assumption between $X$ and $Y$
  
  - Same linear relationship between $X$ and $Y$ for all values of $Z$


---

# Controlling for a variable

- First, let's take `paeduc` and transform it into a categorical measure

```{r}
dat <- dat %>%
  mutate(paeduc_cat = factor(ifelse(paeduc < 9,'8th gr or less',
                             ifelse(paeduc < 12,'HS dropout',
                                    ifelse(paeduc == 12,'HS degree',
                                           ifelse(paeduc < 16,'Some college',
                                                  ifelse(paeduc == 16,'College degree','Post grad'))))),
                             levels = c('8th gr or less','HS dropout','HS degree','Some college','College degree','Post grad')))
dat %>% count(paeduc_cat)
```
---

# Controlling for a variable: Fewest assumptions

```{r,message=F,warning=F}
p <- dat %>%
  ggplot(aes(x = educ,y = realinc)) + 
  geom_point() + 
  geom_smooth() + 
  facet_wrap(~paeduc_cat)
```

---

# Controlling for a variable: Fewest assumptions

```{r,message=F,warning=F}
p
```

---

# Controlling for a variable: More assumptions

```{r}
res <- list()
for(i in unique(dat$paeduc_cat)) {
  res[[i]] <- lm(realinc ~ educ,dat %>% 
                               filter(paeduc_cat == i))
}

# 8th grade or less
summary(res$`8th gr or less`)$coefficients

# High school dropout
summary(res$`HS dropout`)$coefficients
```

---

# Controlling for a variable: More assumptions

```{r}
summary(res$`HS degree`)$coefficients
summary(res$`Some college`)$coefficients
summary(res$`College degree`)$coefficients
summary(res$`Post grad`)$coefficients
```

---

# Controlling for a variable: More assumptions

```{r,echo=F,warning=F,message=F}
toplot <- NULL
for(i in names(res)) {
  toplot <- toplot %>%
    bind_rows(summary(res[[i]])$coefficients %>%
    data.frame() %>%
      slice(2) %>%
    mutate(parent_education = i) %>%
    as_tibble())
}

toplot %>%
  mutate(parent_education = factor(parent_education,levels = c('8th gr or less','HS dropout','HS degree','Some college','College degree','Post grad'))) %>%
  ggplot(aes(x = Estimate,y = parent_education)) + 
  geom_point() + 
  geom_errorbarh(aes(xmin = Estimate - 1.96*`Std..Error`,xmax = Estimate + 1.96*`Std..Error`))
```

---

# Controlling for a variable: Most assumptions

```{r}
summary(lm(realinc ~ educ + paeduc,dat))
```