---
title: "Probability Distributions of $Y$"
subtitle: "and Simulated Sampling Distributions of $\\bar{Y}$"
output: html_document
date: "2023-09-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In the following examples, consider a discrete random variable $Y$ with the probability distribution $p(y)$. As usual $E(Y)= \mu_Y$ and $VAR(Y)= \sigma_Y^2$. The three examples each display tables and graphs illustrating $p(y)$, and then display simulated sampling distributions of $\bar{Y}$  -- the mean of an random sample of $n$ independent observations of the random variable $Y$ -- at sample sizes of $n = 5, 25, and 1000$. The simulations were all constructed using the following process: 

1. Specify some probability distribution $p(y)$.
2. Draw a sample of size $n$ from the probability distribution.
3. Record the sample mean, $\bar{Y}$.
4. Repeat this process 5,000 times.^[Note that I chose 5,000 as a large number that could nevertheless be done in a short amount of time on a standard computer. But this number doesn’t matter. I could have picked 10,000 or 10 million such iterations: at higher numbers of iterations, the histograms would be smoother but would otherwise remain similar.]
5. Display a histogram of the 5,000 $\bar{Y}$’s with 10 bins.

**The take-home point here:** as $N$ becomes large, the Central Limit Theorem tells us that the distribution of the sampling distribution of $\bar{Y}$ converges to the Normal with an ever-smaller variance. This is true, perhaps unsurprisingly, when the distribution of $Y$ is itself nearly Normal (example 1). But it is also true for any and all possible distributions of $Y$, including those that are best described as “bimodal” (example 2) or skewed (example 3). Thus when $n$ is large, no assumptions about the distribution of $Y$ are necessary to fully describe the sampling distribution of $\bar{Y}$ . Under this circumstance, $\bar{Y}$ is distributed Normal with mean $\mu_Y$ and variance $\frac{\sigma_Y^2}{n}$.

## Example 1: The Random Variable $Y$ takes on a nearly **Normal Distribution**

```{r,message=F,warning=F}
set.seed(123)
require(tidyverse)
Y <- rnorm(n = 10000,mean = 5,sd = sqrt(3))
```

### The probability distribution of $Y$ (table)
```{r}
data.frame(Y = round(Y)) %>%
  count(Y) %>%
  mutate(`p(Y=y)` = n / sum(n))
```

### The probability distribution of $Y$ (histogram)
```{r}
data.frame(Y = Y) %>%
  ggplot(aes(x = Y)) + 
  geom_histogram(bins = 20)
```

### The sampling distribution of $\bar{Y}$ at different sizes
```{r}
toplot <- NULL
for(s in c(5,25,1000)) {
  cat(s,'\n')
  for(i in 1:5000) {
    toplot <- toplot %>%
      bind_rows(data.frame(Y = Y) %>%
      sample_n(size = s) %>%
      summarise(ybar = mean(Y)) %>%
        mutate(size = s))
  }
}
 
toplot %>%
  mutate(size = factor(paste0('n = ',size),levels = c('n = 5','n = 25','n = 1000'))) %>%
  ggplot(aes(x = ybar)) + 
  geom_histogram(bins = 60,aes(y = ..density..)) + 
  facet_grid(size~.,scales = 'free_y')
```

Note that we can calculate mean and standard deviation of these results as follows:

```{r}
# Mean: E(Y-bar) = \mu = 5
toplot %>%
  group_by(size) %>%
  summarise(mu = mean(ybar))

# Std Dev: \sigma_Y-bar = sqrt(var(y-bar)) = sqrt(\sigma^2 / n) approx \sqrt(3) / n
toplot %>%
  group_by(size) %>%
  summarise(sigma = sqrt(3/size)) %>%
  distinct()
```

## Example 2: The Random Variable $Y$ takes on a nearly **Bimodal Distribution**

```{r,message=F,warning=F}
set.seed(123)
Y <- c(rnorm(n = 5000,mean = 2,sd = sqrt(2)),rnorm(n = 5000,mean = 7,sd = sqrt(2)))
```

### The probability distribution of $Y$ (table)
```{r}
data.frame(Y = round(Y)) %>%
  count(Y) %>%
  mutate(`p(Y=y)` = n / sum(n))
```

### The probability distribution of $Y$ (histogram)
```{r}
data.frame(Y = Y) %>%
  ggplot(aes(x = Y)) + 
  geom_histogram(bins = 20) + 
  geom_vline(xintercept = mean(Y),linetype = 'dashed') + 
  annotate(geom = 'label',x = mean(Y),y = Inf,label = paste0('mu = ',round(mean(Y),2)),vjust = 1.5)
```

### The sampling distribution of $\bar{Y}$ at different sizes
```{r}
toplot <- NULL
for(s in c(5,25,1000)) {
  cat(s,'\n')
  for(i in 1:5000) {
    toplot <- toplot %>%
      bind_rows(data.frame(Y = Y) %>%
      sample_n(size = s) %>%
      summarise(ybar = mean(Y)) %>%
        mutate(size = s))
  }
}
 
toplot %>%
  mutate(size = factor(paste0('n = ',size),levels = c('n = 5','n = 25','n = 1000'))) %>%
  ggplot(aes(x = ybar)) + 
  geom_histogram(bins = 60,aes(y = ..density..)) + 
  facet_grid(size~.,scales = 'free_y')
```

```{r}
# Mean: E(Y-bar) = \mu = 5
toplot %>%
  group_by(size) %>%
  summarise(mu = mean(ybar))

# Std Dev: \sigma_Y-bar = sqrt(var(y-bar)) = sqrt(\sigma^2 / n) approx \sqrt(3) / n
toplot %>%
  group_by(size) %>%
  summarise(sigma = sqrt(3/size)) %>%
  distinct()
```


## Example 3: The Random Variable $Y$ takes on a **Skewed Distribution**

```{r,message=F,warning=F}
set.seed(123)
Y <- rgamma(n = 10000,shape = 1,rate = 1)
```

### The probability distribution of $Y$ (table)
```{r}
data.frame(Y = round(Y)) %>%
  count(Y) %>%
  mutate(`p(Y=y)` = n / sum(n))
```

### The probability distribution of $Y$ (histogram)
```{r}
data.frame(Y = Y) %>%
  ggplot(aes(x = Y)) + 
  geom_histogram(bins = 20) + 
  geom_vline(xintercept = mean(Y),linetype = 'dashed') + 
  annotate(geom = 'label',x = mean(Y),y = Inf,label = paste0('mu = ',round(mean(Y),2)),vjust = 1.5)
```

### The sampling distribution of $\bar{Y}$ at different sizes
```{r}
toplot <- NULL
for(s in c(5,25,1000)) {
  cat(s,'\n')
  for(i in 1:5000) {
    toplot <- toplot %>%
      bind_rows(data.frame(Y = Y) %>%
      sample_n(size = s) %>%
      summarise(ybar = mean(Y)) %>%
        mutate(size = s))
  }
}
 
toplot %>%
  mutate(size = factor(paste0('n = ',size),levels = c('n = 5','n = 25','n = 1000'))) %>%
  ggplot(aes(x = ybar)) + 
  geom_histogram(bins = 60,aes(y = ..density..)) + 
  facet_grid(size~.,scales = 'free_y')
```

```{r}
# Mean: E(Y-bar) = \mu = 5
toplot %>%
  group_by(size) %>%
  summarise(mu = mean(ybar))

# Std Dev: \sigma_Y-bar = sqrt(var(y-bar)) = sqrt(\sigma^2 / n) approx \sqrt(3) / n
toplot %>%
  group_by(size) %>%
  summarise(sigma = sqrt(3/size)) %>%
  distinct()
```

```{r,message=F,echo=F,warning=F,results='hide'}
dir <- getwd()
type <- 'pdf'
format <- 'portrait'
f <- 'sampling_distributions_of_ybar'

system(paste('Rscript ../../../NFGH/chromeprint.R',dir,type,format,f),wait = F)
```
